Which of my Transient Type Checks are not (Almost) Free?

### 2019-09-12 paper.tex
J's minor tweaks

There is a vast spectrum of different approaches to gradual typing \cite{kafka18,bensurvey18icfp}. Here we measure the performance of ``transient'' or ``type-tag'' checks (as in Reticulated Python), which offer first-order semantics: they check that an object's type constructor or names of supported methods match any static types that the object flows through, but not the return/argument types of such methods \cite{Siek2007,Bloom2009,concrete15,reticPython2014,Greenman2018}.

@@

This has also lead to a technique to evaluate the performance of gradual typing, such as the so called so-called ``Takikawa'' protocol \cite{Takikawa2016,Greenman2019jfp,vitousek-transient-arXive-2019}. The Takikawa protocol was created to measure the run-time cost of gradual typing by testing various configurations of typed/untyped code. This approach was designed to characterise how the amount of typed and untyped code influences performance.
In particular it shows that how simply adding more type annotations does not always produce a uniform effect on performance. Here we adapt their approach in order to identify individual type annotations that may be responsible for performance effects.

@@

Grace allows programmers to choose for each individual declaration whether it should be type-checked, and thus follows languages such as
Reticulated Python \cite{reticPython2014,monotonic2015,Vitousek2017}.
This means in Grace $N$ is the number of type annotations in the program.
This makes it infeasible to checking an entire lattice, even for a moderately sized benchmark.
Vitousek et al.\ therefore modified the Takikawa protocol for
these kinds of languages by taking an approach based on sampling
\cite{vitousek-transient-arXive-2019}.  The Takikawa-Vitousek protocol
divides the number of type annotations in a fully-typed program into
a maximum of 100 intervals, and then randomly generates ten programs within
each interval by erasing type annotations.
However, this approach was designed for benchmarks with large numbers of type annotations, as well as for a larger sample size than our work.

@@

Each benchmark has an almost complete set of type annotations,
\ie, static type annotations are present at almost all possible points.
%ISAAC: I changed this as it was misleading, in a few cases, some of the type annotations are in fact "Unknown", which is exactly what my "removing" of type annotations places

@@

In previous work \cite{roberts-and-co-ecoop-2019}, we also compared the performance of untyped code on Moth against state-of-the-art VMs: Java, Node.js using the V8 JavaScript VM, and the Higgs JavaScript VM. Java was the fastest of these, and on average V8 was about 1.8x slower than Java, Moth was 2.3x slower, and Higgs was 10.4x slower.
We believe, this makes Moth suitable for assessing the impact of type checking,
because Moth's performance is close enough to state-of-the-art VMs,
which avoids hiding of type check overheads in a slow baseline.

@@

By inspecting the scatterplots, we observe that the performance of
Moth in almost any configuration of a benchmark is bounded by
the performance in the untyped and fully-typed configuration.
That is, for these benchmarks on transient type checks,
measuring just the untyped and fully-typed configurations
would provide excellent estimates of a benchmark's
performance bounds.  This is different to the experience
of other kinds of gradual typing, where the best, and most
importantly worst, configurations are not always those
fully typed or fully untyped \cite{Greenman2019jfp}.
%
However, the Richards benchmark \textit{does} have
sections outside these bounds, and isolated executions of
a couple of others (Fannkuch and DeltaBlue) are also outliers.
%
We believe that the 0\% and 100\% bounds are nonetheless a
reasonable heuristic estimator in most cases, and we will
examine further the outlying benchmarks.

@@

The column graphs, right of the corresponding scatter plot, shows the execution time of configurations with only a single type annotation, the x-axis indicates the index of this annotation, thus the first column represents the first annotation, and the last column represents the last one (in the order they appear in the source code). The y-axes for the column graphs are the same as the associated scatter graphs.

@@

As can be seen, most of the time the patterns we found in the typing lattice graphs correspond to outliers in the single type annotation column graphs. For the Snake, Towers and CD benchmarks, there is clearly a pattern where an individual type annotation (highlighted in red) appears responsible for the upper/slowest half of the typing lattice graphs. For Permute there is actually a significant performance \emph{increase}, indicated by the lower half of the lattice being highlighted in red. DeltaBlue and Go also show a slight increase in performance, but here this is caused by two type annotations (in red and blue); this effect is not cumulative, however: the purple dots are about the same height as the red and blue ones, so \emph{either} annotation is sufficient for the full benefit.

@@

In this paper we have investigated how benchmarks can be
repurposed to determine precisely which transient type checks are
likely to cause performance effects,
and for instance are currently not optimised by the just-in-time compiler.
However, detailed analysis will be needed in order to identify exactly what causes such performance effects, such as that we previously did for the List benchmark \cite{roberts-and-co-ecoop-2019}.
We observed that many of our benchmark results
conducted under the Takikawa protocol 
showed a characteristic bimodal performance profile: some
of the benchmark configurations ran significantly slower than
the remaining configurations. We also observed trimodal profiles, as well as performance increases when type annotations where added.

>>>>

There is a vast spectrum of different approaches to gradual typing
\cite{kafka18,bensurvey18icfp}. Here we measure the performance of
``transient'' or ``type-tag'' checks (as in Reticulated Python), which
offer first-order semantics: they check that an object's type
constructor or names of supported methods match any static types that
the object flows through, but not the return types or argument types of those methods \cite{Siek2007,Bloom2009,concrete15,reticPython2014,Greenman2018}.

@@

This has also lead to a technique to evaluate the performance of gradual typing, the ``Takikawa'' protocol \cite{Takikawa2016,Greenman2019jfp,vitousek-transient-arXive-2019}. The Takikawa protocol was created to measure the run-time cost of gradual typing by testing various configurations of typed and untyped code. This approach was designed to characterise how the amount of typed and untyped code influences performance.
In particular, Takikawa protocol evaluations often show that how simply adding more type annotations does not always produce a uniform effect on performance. Here we adapt their approach in order to identify individual type annotations that may be responsible for performance effects.

@@

Grace allows programmers to choose whether each individual declaration should be type-checked, and thus follows languages such as
Reticulated Python \cite{reticPython2014,monotonic2015,Vitousek2017}.
This means in Grace $N$ is the number of type annotations in the program, so it is
infeasible to check an entire lattice, even for a moderately sized benchmark.
Vitousek et al.\ therefore modified the Takikawa protocol for
these kinds of languages by taking an approach based on sampling
\cite{vitousek-transient-arXive-2019}.  The Takikawa-Vitousek protocol
divides the number of type annotations in a fully-typed program into
a maximum of 100 intervals, and then randomly generates ten programs within
each interval by erasing type annotations.
However, this approach was designed for benchmarks with large numbers of type annotations, as well as for a larger sample size than our work.

@@

%Each benchmark has an almost complete set of type annotations,
%\ie, static type annotations are present at almost all possible points.
%ISAAC: I changed this as it was misleading, in a few cases, some of the type annotations are in fact "Unknown", which is exactly what my "removing" of type annotations places

We added type annotations to every benchmark, aiming to use the most appropriate (most specific) annotation for each declaration.
%KJX: I changed it too because I though that was also misleading.

@@

In previous work \cite{roberts-and-co-ecoop-2019}, we also compared the performance of untyped code on Moth against state-of-the-art VMs: Java, Node.js using the V8 JavaScript VM, and the Higgs JavaScript VM. Java was the fastest of these, and on average V8 was about 1.8x slower than Java, Moth was 2.3x slower, and Higgs was 10.4x slower.
We believe this makes Moth suitable for assessing the impact of type checking,
because Moth's performance is close enough to state-of-the-art VMs,
which should make it harder to hide type checking overheads in a slow baseline.

@@

By inspecting the scatterplots, we observe that the performance of
Moth in almost any configuration of a benchmark is bounded by
the performance in the untyped and fully-typed configuration.
That is, for these benchmarks on transient type checks,
measuring just the untyped and fully-typed configurations
would provide excellent estimates of a benchmark's
performance bounds.  This is different to the experience
of other kinds of gradual typing, where the best, and most
importantly worst, configurations are not always those
fully typed or fully untyped \cite{Greenman2019jfp}.
%
However, the Richards benchmark \textit{does} have
sections outside these bounds, and isolated executions of
a couple of others (Fannkuch and DeltaBlue) are also outliers.
%
We believe that the 0\% and 100\% bounds are nonetheless a
reasonable heuristic estimator in most cases, and we will
examine the outlying benchmarks further.

@@

The column graphs, to the right of the corresponding scatter plot, show the execution time of configurations with only a single type annotation, the x-axis indicates the index of this annotation, thus the first column represents the first annotation, and the last column represents the last one (in the order they appear in the source code). The y-axes for the column graphs are the same as the associated scatter graphs.

@@

As can be seen, most of the patterns we found in the typing lattice graphs correspond to outliers in the single type annotation column graphs. For the Snake, Towers and CD benchmarks, there is clearly a pattern where an individual type annotation (highlighted in red) appears responsible for the upper/slowest half of the typing lattice graphs. For Permute there is actually a significant performance \emph{increase}, indicated by the lower half of the lattice being highlighted in red. DeltaBlue and Go also show a slight increase in performance, but here this is caused by two type annotations (in red and blue); this effect is not cumulative, however: the purple dots are about the same height as the red and blue ones, so \emph{either} annotation is sufficient for the full benefit.

@@

In this paper we have investigated how benchmarks can be
repurposed to determine precisely which transient type checks are
likely to cause performance effects,
and for instance are currently not optimised by the just-in-time compiler.
However, detailed analysis will be needed in order to identify exactly what causes such performance effects: we previously undertook such an analysis for the List benchmark \cite{roberts-and-co-ecoop-2019}.

We observed that many of our benchmark results
conducted under the Takikawa protocol 
showed a characteristic bimodal performance profile: some
of the benchmark configurations ran significantly slower than
the remaining configurations. We also observed trimodal profiles, as well as performance increases when type annotations where added.
